---
title: "Brms hacking: linear predictors for random effect standard deviations"
date: 2024-02-17
draft: true
tags: ["R","Stan","brms", "random effects"]
---

`brms` is a great package. It allows you to put predictors on a lot of things.
Its power is however not absolute --- one thing it doesn't let you directly do is use data to predict variances of random/varying effects. Here we will show pretty general techniques
to hack with `brms` that let us achieve exactly this goal (and many more).

To be precise, you can use the construct `(1|gr(patient, by = trt))` which fits a 
separate standard deviation for each level of `trt`, which
is almost the same as using `trt` as a categorical predictor for the standard deviation. 
You however cannot go further and use any other type of predictors here. E.g. the following model is impossible in plain `brms` :

$$
y_i \sim N\left(\mu_i, \sigma \right) \\
\mu_i = \alpha + \beta x_i + \gamma_{\text{patient}(i)} \\
\gamma_{p} \sim N \left(0, \tau_{\text{treatment}(p)}\right) \\
\tau_t = \alpha^\prime + \beta^\prime x^\prime_t
$$

Where $x$ is a vector of observation-level predictors while $x^\prime$ is a vector of treatment-level predictors. 
In between we have patients --- each contributing a bunch of observations and the standard deviation of the 
patient-level random intercepts depends on our treatment-level predictors.


Well, it is not completely impossible. Since `brms` is _immensely_ hackable, you can
actually make this work. This blogpost will discuss how to do this. This does not
mean it is a good idea or that you _should_ do it. I am just showing that it is 
possible and hopefully also showing some general ways to hack with `brms`.

The main downside of my approach is that it forces you to completely override
the likelihood and that you have to build the random effect with predicted sigma
in hand-written Stan code. This may mean the benefits of `brms` are now too small and 
you might be better off building the whole thing directly in Stan.

Also, this type of model is likely to be a bit data-hungry --- you need to have
enough observations per treatment and enough treatments to be able to estimate $\tau$ 
well enough to learn about its predictors.

## Setting up

Let's set up and get our hands dirty.

```{r setup, message=FALSE, warning=FALSE}
library(cmdstanr)
library(brms)
library(tidyverse)
library(knitr)
library(bayesplot)

ggplot2::theme_set(cowplot::theme_cowplot())
options(mc.cores = parallel::detectCores(), brms.backend = "cmdstanr")

cache_dir <- "_brms_ranef_cache"
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

```


## Simulate data

Note that the way we have setup the model implies that patients are nested within treatments (i.e. that each patient only ever gets one treatment). 
Since each random effect can only have one prior distribution, this is the easiest way to make sense of the model.

_Note: Another interpretation of "predicting random effect sd" is that you take an $N(0,1)$ distributed random effect and then multiply it by a different value based on other predictors. It should be possible to achieve that with
non-linear formulas in `brms` and this blog does not address this case_.

First, we setup the treatment-level predictors in a treatment-level data frame and use those to predict the sds ($\tau$ above).

```{r}
set.seed(354855)
N <- 500
N_pts <- floor(N / 5)
N_trts <- 10
trt_intercept <- 0
trt_x_b <- 1
trt_data <- data.frame(trt_x = rnorm(N_trts))
# Corresponds to tau in the mathematical model
trt_sd <- exp(trt_intercept + trt_x_b * trt_data$trt_x)
```

Now, we setup the patient-level random effects, with varying sds (corresponding to $\gamma$ above).

```{r}
patient_treatment <- sample(1:N_trts, size = N_pts, replace = TRUE)
ranef <- rnorm(N_pts, mean = 0, sd = trt_sd[patient_treatment])
```

Finally, we setup the main data frame with multiple observations of each patient.

```{r}
intercept <- 1
x_b <- 0.5
obs_sigma <- 1
base_data <- data.frame(x = rnorm(N), 
                        patient_id = rep(1:N_pts, length.out = N))

base_data$trt_id <- patient_treatment[base_data$patient_id]

base_data_predictor <- intercept + x_b * base_data$x + ranef[base_data$patient_id]
base_data$y <- rnorm(N, mean = base_data_predictor , sd = obs_sigma)
```


## Implement the model

The first problem to solve is that at its core, `brms` requires us to use a single data
frame as input. But we have a treatment-level data frame and then an observation-level 
data frame. We get around this by adding dummy values so that both data 
frames have the same columns, binding the together and then using the `subset`
addition term to use different formulas for each. We will also need a dummy outcome
variable for the treatment-level data.

```{r}
combined_data <- rbind(
  base_data %>% mutate(
    is_trt = FALSE,
    trt_x = 0,
    trt_y = 0
  ),
  trt_data %>% mutate(
    is_trt = TRUE,
    trt_id = 0,
    patient_id = 0,
    y = 0,
    x = 0,
    trt_y = 0
  )
)
```


The main idea for implementation is that we completely overtake the machinery of `brms` after
the linear predictors are constructed. To do that, we create a custom family
that is empty (i.e. adds nothing to the log likelihood) and use it in our formula.

```{r}
# Build the empty families --- one has just a single parameter and will be used 
# for treatment-level sds. The other has mu and sigma parameter will be used
# for the observation model.
empty_for_trt <- custom_family("empty_for_trt", type = "real")
empty_for_obs <- custom_family("empty_for_obs",  dpars = c("mu", "sigma"), 
                               links = c("identity", "log"), type = "real", lb = c(NA, 0))

empty_func_stanvar <- stanvar(block = "functions", scode = "
  real empty_for_trt_lpdf(real y, real mu) {
    return 0;
  } 
  
  real empty_for_obs_lpdf(real y, real mu, real sigma) {
    return 0;
  }
")

```


We then take the linear predictions for the sd of the random effects ($\tau$) 
and use it to
manually build our random effect values (with non-centered parametrization). 
We manually add those values to the
rest of the linear predictor term and then manually add our desired likelihood.

This will let our final formula to look this way:

```{r}
f <- mvbrmsformula(
    brmsformula(y | subset(!is_trt)  ~ x, family = empty_for_obs),
    brmsformula(trt_y | subset(is_trt)  ~ trt_x, family = empty_for_trt),
    rescor = FALSE)

```


In this setup, `brms` will build a bunch of variables for both formulas
that we can access in our Stan code.
Their names will depend on the name of the outcome variables --- since our
main outcome is `y`, relevant variables will be `N_y` (number of rows
for this outcome), `mu_y` and `sigma_y` (distributional parameters for this outcome).

Our dummy outcome is `trt_y` and the relevant variables will be `N_trty` and `mu_trty`,
because `brms` removes underscores. You can always use `make_stancode` and
`make_standata` to see how `brms` transforms names and input data.

For all this to happen we also need to pass a bunch of extra data via `stanvars`.

Let us prepare the extra Stan code.

```{r}
# Pass the extra data. We'll take advantage of some already existing data
# variables defined by brms, these include:
# N_y - the number of observation-level data 
# N_trty - the number of treatment-level data (and thus the number of treatments)
# we however need to pass the rest of the data for the random effect
data_stanvars <- 
  stanvar(x = N_pts, block = "data", scode = "int<lower=2> N_pts;") +
  stanvar(x = patient_treatment, name = "trt_id", block = "data", 
          scode = "array[N_pts] int<lower=1, upper=N_trty> trt_id;") +
  stanvar(x = base_data$patient_id, name = "patient_id", block = "data", 
          scode = "array[N_y] int<lower=1, upper=N_pts> patient_id;")

# Raw parameters for the random effects
parameter_stanvar <- 
  stanvar(block = "parameters", scode = "
      vector[N_pts] my_ranef_raw;
      ")

# Prior - we are using the non-centered parametrization, so it is just N(0,1)
# and we multiply by the sd later.
# Note that current versions of brms compute log-prior in the transformed
# parameters block, so we do it as well.
prior_stanvar <-
  stanvar(block = "tparameters", 
          scode = "lprior += std_normal_lpdf(to_vector(my_ranef_raw));")

# Here is where we add the random effect to the existing predictor values and 
# reconstruct the likelihood.
# Once again using the values generated by brms for the predictors in mu_trty,
# mu_y, sigma_y.
likelihood_stanvar <- 
  stanvar(block = "likelihood", position = "end", scode = "
      // New scope to let us introduce new parameters
      {
          vector[N_trty] trt_sds = exp(mu_trty);
          vector[N_pts] my_ranef = my_ranef_raw .* trt_sds[trt_id];
          for(n in 1: N_y) {
            // Add the needed ranef 
            real mu_with_ranef = mu_y[n] + my_ranef[patient_id[n]];
            // reimplement the likelihood
            target += normal_lpdf(Y_y[n] | mu_with_ranef, sigma_y);
          }
      }
          ") 
  

predict_ranef_stanvars <- empty_func_stanvar + 
  data_stanvars + 
  parameter_stanvar + 
  prior_stanvar +
  likelihood_stanvar
```

This is the complete Stan code generated by `brms` with our additions:

```{r}
make_stancode(f, data = combined_data,
  stanvars = predict_ranef_stanvars)
```

Now, we can compile and fit the model:

```{r}
fit <- brm(  
  f, 
  data = combined_data,
  stanvars = predict_ranef_stanvars,
  file = file.path(cache_dir, "fit.rds"),
  file_refit = "on_change")
```

We get a decent recovery of the parameters --- recall that we simulated data with
`y_Intercept` = `r intercept`, `trty_Intercept` = `r trt_intercept`, `y_x` = `r x_b`,
`trty_trt_x` = `r trt_x_b` and `sigma_y` = `r obs_sigma`.

```{r}
summary(fit)
```

Unfortunately, the code above is somewhat fragile. Notably, if we add a predictor
for the standard deviation of the observations, then `sigma_y` in the Stan code
won't be a scalar, but a vector and we'll need to adjust the Stan code a little bit.

## Using the fitted model

Since we overtook so much of `brms` machinery, things like `posterior_predict()`, 
 `posterior_epred()` and `log_lik()` won't work out of the box and we need a little extra work to
get them, mirroring the extra steps we did in the Stan code.

Luckily for us `brms` exposes the [`prepare_predictions()`](http://paul-buerkner.github.io/brms/reference/prepare_predictions.html)
and [`get_dpar()`](http://paul-buerkner.github.io/brms/reference/get_dpar.html) functions
which do most of the heavy lifting. Let's start with mimicking `posterior_epred()` 

```{r}
pred_trt <- prepare_predictions(fit, resp = "trty")
# A matrix of 4000 draws per 10 treatments
samples_trt_mu <- brms::get_dpar(pred_trt, "mu")

pred_y <- prepare_predictions(fit, resp = "y")
# A matrix of 4000 draws per 500 observations
samples_mu <- brms::get_dpar(pred_y, "mu") 

# the ranef samples need to be taken directly from the Stan fit
# A matrix of 4000 draws per 100 patients
samples_ranef_raw <- posterior::as_draws_matrix(fit$fit) %>% 
  posterior::subset_draws(variable = "my_ranef_raw")

samples_sigma_per_patient <- exp(samples_trt_mu)[, patient_treatment]
samples_ranef <- samples_ranef_raw * samples_sigma_per_patient
samples_ranef_per_obs <- samples_ranef[, base_data$patient_id]
samples_epred <- samples_mu + samples_ranef_per_obs 
```

And once we have the predictions for `mu` we can combine them with samples
for `sigma` to get predictions including the observation noise and continue
to do a posterior predictive check (which looks good).

```{r ppc}
# A vector of 4000 draws
samples_sigma <- brms::get_dpar(pred_y, "sigma") 


pred_y <- matrix(nrow = nrow(samples_epred), ncol = ncol(samples_epred))
for(j in 1:ncol(samples_epred)) {
  pred_y[,j] <- rnorm(nrow(samples_epred), 
                      mean = samples_epred[,j], 
                      sd = samples_sigma)
}

bayesplot::ppc_dens_overlay(base_data$y, pred_y[sample.int(4000, size = 30),])

```

## Summary

So yay, we can use `brms` for the core of our model and then extend it to 
cover predictors for the standard deviation of random effects. Unfortunately,
it requires quite a bit of extra work.
Using this heavy machinery for such a simple model as we did in this quick
example is probably an overkill
and you would be better off just implementing the whole thing in Stan. But if your 
current `brms` model is quite complex and the only extra thing you need are the
sd predictors, then the cost-benefit considerations might be quite different.

The techniques we used to hack around `brms` are also very general, note that
we have shown how to:

- Combine multiple datasets of different sizes/shapes in a single model
- Replace likelihood with arbitrary Stan code

Together this is enough to use `brms`-style predictors in connection with 
basically any type of model.
For example, these tricks power my implementation of hidden Markov models with `brms` 
discussed at https://discourse.mc-stan.org/t/fitting-hmms-with-time-varying-transition-matrices-using-brms-a-prototype/19645/7 .

## Appendix: Check with SBC

Recovering parameters from a single simulation and a nice posterior predictive
check are good starting points but far from a guarantee that we implemented the
model correctly. To be sure, we'll check with SBC --- if you are not familiar,
SBC is a method that can discover almost all implementation problems  in your model
by repeatedly fitting simulated data.
We'll use the [`SBC` R package](https://hyunjimoon.github.io/SBC) and won't
explain all the details here --- check the [Getting started](https://hyunjimoon.github.io/SBC/articles/SBC.html) and [SBC for `brms`](https://hyunjimoon.github.io/SBC/articles/brms.html) vignettes for
explanation of the main concepts and API.

```{r}
# Setting up SBC and paralellism
library(SBC)
future::plan(future::multisession)
gamma_shape <- 14
gamma_rate <- 4
trt_intercept_prior_mu <- 0.5
trt_intercept_prior_sigma <- 0.5
```


To make the model work with SBC we add explicit priors for all parameters (as the simulations need to match
those priors). We'll use $N(0,1)$ for most parameters except the intercept for
random effect deviations ($\alpha^\prime$) where we'll use $N(`r trt_intercept_prior_mu`,`r trt_intercept_prior_sigma`)$ to avoid both very low 
and very large standard deviations which pose convergence problems. Similarly, very low observation sigma
causes convergence problems, so we'll use a $\Gamma(`r gamma_shape`, `r gamma_rate`)$ prior (roughly saying that a priori
the standard deviation is unlikely to be less than `r round(qgamma(0.025, gamma_shape, gamma_rate), 1)` or
more than `r round(qgamma(0.975, gamma_shape, gamma_rate), 1)` ). I did not investigate deeply
to understand the convergence issues, so not completely sure about the mechanism.

```{r, message=FALSE, warning=FALSE}
get_prior(f, combined_data)

priors <- c(
  set_prior("normal(0,1)", class = "b", resp = "trty"),
  set_prior(paste0("normal(",trt_intercept_prior_mu, ", ", 
                   trt_intercept_prior_sigma, ")"), 
            class = "Intercept", resp = "trty"),
  set_prior("normal(0,1)", class = "b", resp = "y"),
  set_prior("normal(0,1)", class = "Intercept", resp = "y"),
  set_prior(paste0("gamma(", gamma_shape, ", ", gamma_rate, ")"), 
            class = "sigma", resp = "y")
)

# Function to generate a single simulated dataset
# Note: we reuse N_trts, N, N_pts, patient_id and patient_treatment 
# from the previous code to keep the data passed via stanvars fixed.
generator_func <- function() {
  trt_intercept <- rnorm(1, mean = trt_intercept_prior_mu, 
                         sd = trt_intercept_prior_sigma)
  trt_x_b <- rnorm(1)

  trt_data <- data.frame(trt_x = rnorm(N_trts))
  # Centering predictors to match brms
  trt_data$trt_x <- trt_data$trt_x - mean(trt_data$trt_x)
  trt_sd <- exp(trt_intercept + trt_x_b * trt_data$trt_x)
  
  ranef_raw <- rnorm(N_pts)
  ranef <- ranef_raw * trt_sd[patient_treatment]
  
  intercept <- rnorm(1)
  x_b <- rnorm(1)
  obs_sigma <- rgamma(1, gamma_shape, gamma_rate)
  
  obs_data <- data.frame(x = rnorm(N), 
                          patient_id = base_data$patient_id)
  
  obs_data$x <- obs_data$x - mean(obs_data$x)
  obs_data$trt_id <- patient_treatment[obs_data$patient_id]
  
  obs_data_predictor <- intercept + x_b * obs_data$x + ranef[obs_data$patient_id]
  obs_data$y <- rnorm(N, mean = obs_data_predictor , sd = obs_sigma)
  
  combined_data <- rbind(
    obs_data %>% mutate(
      is_trt = FALSE,
      trt_x = 0,
      trt_y = 0
    ),
    trt_data %>% mutate(
      is_trt = TRUE,
      trt_id = 0,
      patient_id = 0,
      y = 0,
      x = 0,
      trt_y = 0
    )
  )
  
  list(generated = combined_data,
       variables = list(
         b_y_Intercept = intercept,
         b_y_x = x_b,
         b_trty_Intercept = trt_intercept,
         b_trty_trt_x = trt_x_b,
         sigma_y = obs_sigma,
         my_ranef_raw = ranef_raw
       ))
}

# Generate a lot of datsets
set.seed(33214855)
N_sims <- 1000
ds <- generate_datasets(SBC_generator_function(generator_func), n_sims = N_sims)

# With 1000 datasets, this takes ~45 minutes on my computer
backend <-
  SBC_backend_brms(f,
                   stanvars = predict_ranef_stanvars,
                   prior = priors,
                   template_data = combined_data,
                   chains = 2,
                   out_stan_file = file.path(cache_dir, "backend.stan")
                   )
```


To increase the power of SBC to detect problems, we will also add the 
log-likelihood and log-prior as derived quantities (see [Modrák et al. 2023](https://doi.org/10.1214/23-BA1404) 
or the [limits of SBC](https://hyunjimoon.github.io/SBC/articles/limits_of_SBC.html) vignette for background on this). 

```{r}
compute_loglik <- function(y, is_trt, x, trt_x, patient_id, 
                           trt_id, intercept, x_b, trt_intercept, trt_x_b, 
                           ranef_raw, sigma_y) {
  patient_id <- patient_id[!is_trt]
  trt_id <- trt_id[!is_trt]
  x <- x[!is_trt]
  y <- y[!is_trt]
  
  trt_x <- trt_x[is_trt]
  
  patient_trt_all <- matrix(nrow = length(patient_id), ncol = 2)
  patient_trt_all[,1] <- patient_id
  patient_trt_all[,2] <- trt_id
  patient_trt_all <- unique(patient_trt_all)

  patient_treatment <- integer(max(patient_id))
  patient_treatment[patient_trt_all[, 1]] <- patient_trt_all[, 2]
  
  ranef_sigma <- exp(trt_intercept + trt_x * trt_x_b)
  ranef_vals <- ranef_raw * ranef_sigma[patient_treatment]
  mu <- intercept + x * x_b + ranef_vals[patient_id]
  sum(dnorm(y, mean = mu, sd = sigma_y, log = TRUE))  
}


dq <- derived_quantities(
  lprior_fixed = dnorm(b_y_Intercept, log = TRUE) +
    dnorm(b_y_x, log = TRUE) +
    dnorm(b_trty_Intercept, mean = trt_intercept_prior_mu, 
          sd = trt_intercept_prior_sigma, log = TRUE) +
    dnorm(b_trty_trt_x, log = TRUE) +
    dgamma(sigma_y, gamma_shape, gamma_rate, log = TRUE),
  loglik = compute_loglik(y = y, is_trt = is_trt, x = x, trt_x = trt_x,
                          patient_id = patient_id, trt_id = trt_id,
                          intercept = b_y_Intercept, x_b = b_y_x,
                          trt_intercept = b_trty_Intercept, 
                          trt_x_b = b_trty_trt_x, ranef_raw = my_ranef_raw,
                          sigma = sigma_y),
  .globals = c("compute_loglik", "gamma_shape", "gamma_rate",
               "trt_intercept_prior_mu", "trt_intercept_prior_sigma")
)
```


We are now ready to actually run SBC:

```{r}
sbc_res <-
  compute_SBC(
    ds,
    backend,
    dquants = dq,
    cache_mode = "results",
    cache_location = file.path(cache_dir, paste0("sbc", N_sims, ".rds")),
    keep_fits = N_sims <= 50
  )
```

There are still some convergence problems for some fits --- the most worrying are 
the high Rhats, affecting almost a third of the fits. This should definitely
warrant some further investigation, but the Rhats are not very large and this 
is a blog post, not a research paper,
so we will not go down this rabbit hole.

A very small number of fits had divergences/treedepth issues, but due to the small
number, those are not so worrying. The steps rejected is completely benign
as this includes rejections during warmup.

Overall, it is in fact safe to just ignore the problematic fits as long as 
you would not use results from such fits in actual practice (which you shouldn't) ---
see the [rejection sampling](https://hyunjimoon.github.io/SBC/articles/rejection_sampling.html) vignette for more details.

We plot the results of the [ECDF diff check](https://hyunjimoon.github.io/SBC/articles/rank_visualizations.html#plot_ecdf-and-plot_ecdf_diff---ecdf-plots) --- looking good!


```{r sbc-ecdf}
vars <- sbc_res$stats %>% filter(!grepl("my_", variable)) %>% 
  pull(variable) %>% unique() %>% c("my_ranef_raw[1]", "my_ranef_raw[2]")

excluded_fits <- sbc_res$backend_diagnostics$n_divergent > 0 |
  sbc_res$backend_diagnostics$n_max_treedepth > 0 |
  sbc_res$default_diagnostics$min_ess_tail < 200 |
  sbc_res$default_diagnostics$max_rhat > 1.01
sbc_res_filtered <- sbc_res[!excluded_fits]
plot_ecdf_diff(sbc_res_filtered, variables = vars)
```


We can also see how close to the true values our estimates are --- once again this looks
quite good --- we do learn quite a lot of information about all parameters except for the 
random effects!

```{r sbc-sim-est}
plot_sim_estimated(sbc_res_filtered, variables = vars, alpha = 0.1)
```

And that's all. If you encounter problems running the models that you can't resolve yourself, be
sure to ask questions on [Stan Discourse](https://discourse.mc-stan.org) and tag
me (@martinmodrak) in the question!

## Original computing environment 

```{r, cache=FALSE, echo=FALSE}
git_rev <- tryCatch({system("git rev-parse HEAD", intern=TRUE)}, error = function(e) {"Could not read Git revision"})
```

This post was built from Git revision [``r git_rev``](https://github.com/martinmodrak/blog/tree/`r git_rev`), you can download the [`renv.lock`](https://github.com/martinmodrak/blog/blob/`r git_rev`/renv.lock) file required to reconstruct the environment.


```{r}
sessionInfo()
```

