---
title: "Putting predictors on random effect standard deviations with brms"
date: 2021-04-01
draft: true
tags: ["R","Stan","brms", "random effects"]
---

`brms` is a great package. It allows you to put predictors on a lot of things.
Its power is however not absolute --- one thing it doesn't let you do is use data to predict variances of random/varying effects. 

To be precise, you can use the construct `(1|gr(patient, by = Trt))` which fits a 
separate standard deviation for each level of `Trt`, which
is almost the same as using `Trt` as a categorical predictor for the standard deviation. 
You however cannot go further and use more complex predictors here. E.g. the following model is impossible in plain `brms` :

$$
y_i \sim N\left(\mu_i, \sigma \right) \\
\mu_i = \alpha + \beta x_i + \gamma_{\text{patient}(i)} \\
\gamma_{p} \sim N \left(0, \tau_{\text{treatment}(p)}\right) \\
\tau_t = \alpha^\prime + \beta^\prime x^\prime_t
$$

Where $x$ is a vector of observation-level predictors while $x^\prime$ is a vector of treatment-level predictors. 
In between we have patients --- each contribution a bunch of observations and the standard deviation of the 
patient-level random intercepts depends on our treatment-level predictors.


Well, it is not completely impossible to build such a model. Since `brms` is _immensely_ hackable, you can
actually make this work. This blogpost will discuss how to do this. This does not
mean it is a good idea or that you _should_ do it. I am just showing that it is 
possible and hopefully also showing more general ways to hack with `brms`.

The main downside of my approach is that it forces you to completely override
the likelihood and that you have to build the random effect with predicted sigma
manually in Stan. This may mean the benefits of `brms` are now too small and 
you might be better off building the whole thing directly in Stan.

## Setting up

Let's set up and get our hands dirty.

```{r setup, message=FALSE, warning=FALSE}
library(cmdstanr)
library(brms)
library(tidyverse)
library(knitr)
library(bayesplot)

ggplot2::theme_set(cowplot::theme_cowplot())
options(mc.cores = parallel::detectCores(), brms.backend = "cmdstanr")

cache_dir <- "_brms_ranef_cache"
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

```


## Simulate data

Note that the way we have setup the model implies that patients are nested within treatments (i.e. that each patient only ever gets one treatment). 
Since each random effect can only have one prior distribution, this is the easiest way to make sense of the model.

_Note: Another interpretation of "predicting random effect sd" is that you take a random effect and then multiply it by a different value based on other predictors. This should be possible to achieve that with
non-linear formulas and this blog does not address this case_.

First, we setup the treatment-level predictors in a treatment-level data frame and use those to predict the sds ($\tau$ above).

```{r}
set.seed(354855)
N <- 1000
N_pts <- floor(N / 3)
N_trts <- 10
trt_intercept <- 0
trt_x_b <- 1
trt_data <- data.frame(trt_x = rnorm(N_trts))
# Corresponds to tau in the mathematical model
trt_sd <- exp(trt_intercept + trt_x_b * trt_data$trt_x)
```

Now we setup the patient-level random effects, with varying sds (corresponding to $\gamma$ above).

```{r}
patient_treatment <- sample(1:N_trts, size = N_pts, replace = TRUE)
ranef <- rnorm(N_pts, mean = 0, sd = trt_sd[patient_treatment])
```

Now we setup the main data frame showing multiple observations of each patient.

```{r}
intercept <- 1
x_b <- 0.5
obs_sigma <- 1
base_data <- data.frame(x = rnorm(N), 
                        patient_id = rep(1:N_pts, length.out = N))

base_data$trt_id <- patient_treatment[base_data$patient_id]

base_data_predictor <- intercept + x_b * base_data$x + ranef[base_data$patient_id]
base_data$y <- rnorm(N, mean = base_data_predictor , sd = obs_sigma)
```


## Implement the model

The first problem to solve is that at its core, `brms` requires us to use a single data
frame as input, but we have a treatment-level data frame and then an observation-level 
data frame. We get around this by adding dummy values so that both data 
frames have the same column, binding the together and then using the `subset`
addition term use different formulas for each. We will also need a dummy outcome
variable for the treatment-level data.

```{r}
combined_data <- rbind(
  base_data %>% mutate(
    is_trt = FALSE,
    trt_x = 0,
    trt_y = 0
  ),
  trt_data %>% mutate(
    is_trt = TRUE,
    trt_id = 0,
    patient_id = 0,
    y = 0,
    x = 0,
    trt_y = 0
  )
)
```


The main idea for implementation is that we completely overtake the machinery of `brms` after
the linear predictors are constructed. To do that, we create a custom family
that is empty (i.e. adds nothing to the log likelihood) and use it.

```{r}
# Build the empty families --- one has just a single parameter and will be used 
# for treatment-level sds. The other has mu and sigma parameter will be used
# for the observation model.
empty_for_trt <- custom_family("empty_for_trt", type = "real")
empty_for_obs <- custom_family("empty_for_obs",  dpars = c("mu", "sigma"), 
                               links = c("identity", "log"), type = "real")

empty_func_stanvar <- stanvar(block = "functions", scode = "
  real empty_for_trt_lpdf(real y, real mu) {
    return 0;
  } 
  
  real empty_for_obs_lpdf(real y, real mu, real sigma) {
    return 0;
  }
")

```


We then take the linear predictions for the sd of the random effects ($\tau$) 
and use it to
manually build our random effect values (with non-centered parametrization). 
We manually add those values to the
rest of the linear predictor term and then manually add our desired likelihood.

This will let our final formula to look this way:

```{r}
f <- mvbrmsformula(
    brmsformula(y | subset(!is_trt)  ~ x, family = empty_for_obs),
    brmsformula(trt_y | subset(is_trt)  ~ trt_x, family = empty_for_trt),
    rescor = FALSE)

```


In this setup, `brms` will build a bunch of variables for both formulas.
Their names will depend on the name of the outcome variables --- since our
main outcome is `y`, relevant variables will be `N_y` (number of rows
for this outcome), `mu_y` and `sigma_y` (distributional parameters for this outcome).

Our dummy outcome is `trt_y` and the relevant variables will be `N_trty` and `mu_trty`,
because `brms` removes underscores. You can always use `make_stancode` and
`make_standata` to see how `brms` transforms names and input data.

For all this to happen we also need to pass a bunch of extra data.

Let us prepare the extra Stan code.

```{r}
# Pass the extra data. We'll take advantage of some already existing data
# variables defined by brms, these include:
# N_y - the number of observation-level data 
# N_trty - the number of treatment-level data (and thus the number of treatments)
data_stanvars <- 
  stanvar(x = N_pts, block = "data", scode = "int<lower=2> N_pts;") +
  stanvar(x = patient_treatment, name = "trt_id", block = "data", 
          scode = "array[N_pts] int<lower=1, upper=N_trty> trt_id;") +
  stanvar(x = base_data$patient_id, name = "patient_id", block = "data", 
          scode = "array[N_y] int<lower=1, upper=N_pts> patient_id;")

# Raw parameters for the random effects
parameter_stanvar <- 
  stanvar(block = "parameters", scode = "
      vector[N_pts] my_ranef_raw;
      ")

# Prior - we are using non-centered parametrization, so it is just N(0,1)
# and we multiply by the sd later.
# Note that current versions of brms compute log-prior in the transformed
# parameters block, so we do it as well.
prior_stanvar <-
  stanvar(block = "tparameters", 
          scode = "lprior += std_normal_lpdf(to_vector(my_ranef_raw));")

# Here is where we add the random effect to the existing predictor values and 
# reconstruct the likelihood.
# Once again using the values generated by brms for the predictors in mu_trty,
# mu_y, sigma_y.
likelihood_stanvar <- 
  stanvar(block = "likelihood", position = "end", scode = "
      // New scope to let us introduce new parameters
      {
          vector[N_trty] trt_sds = exp(mu_trty);
          vector[N_pts] my_ranef = my_ranef_raw .* trt_sds[trt_id];
          for(n in 1: N_y) {
            // Add the needed ranef 
            real mu_with_ranef = mu_y[n] + my_ranef[patient_id[n]];
            // reimplement the likelihood
            target += normal_lpdf(Y_y[n] | mu_with_ranef, sigma_y);
          }
      }
          ") 
  

predict_ranef_stanvars <- empty_func_stanvar + 
  data_stanvars + 
  parameter_stanvar + 
  prior_stanvar +
  likelihood_stanvar
```

This is the complete Stan code generated by `brms` with our additions:

```{r}
make_stancode(f, data = combined_data,
  stanvars = predict_ranef_stanvars)
```


```{r, include=FALSE}
dd <- make_standata(  
  f, 
  data = combined_data,
  stanvars = predict_ranef_stanvars)

dd$N_y
```


Now, we can compile and fit the model:

```{r}
fit <- brm(  
  f, 
  data = combined_data,
  stanvars = predict_ranef_stanvars)
```



```{r}
summary(fit)
```



If you encounter problems running the models that you can't resolve yourself, be
sure to ask questions on [Stan Discourse](https://discourse.mc-stan.org) and tag
me (@martinmodrak) in the question!

## Original computing environment 

```{r}
sessionInfo()
```

